Investigation of scarid data

### 2018-11-13 r/Data_prodding.Rmd
11/12
11/13

---
title: "Scarid grazing analysis"
author: "Molly Wilson"
date: "Dec. 2018"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    number_sections: true

@@

```{r,message=FALSE,warning=FALSE, echo=FALSE}
# set up
library(startR)
library(here)
library(tidyverse)
library(RColorBrewer)
library(corrplot)
library(nlme)
library(pls)
library(factoextra)
```
```{r,message=FALSE,warning=FALSE, echo=FALSE}
sum_site <- read.csv(here("data","sum_site.csv"), stringsAsFactors=F) %>%
  select(-X)
sum_id <- read.csv(here("data","sum_id.csv"), stringsAsFactors=F) %>%
  select(-X)

# stop<-sum_id%>%filter(species_code=="stop")
# qup<-sum_id%>%filter(species_code=="qup")
# rbp<-sum_id%>%filter(species_code=="rbp")
```

@@

```{r,echo=FALSE}
sample <- group_by(sum_id, site, species, phase) %>% summarize(n=n())
barplot(table(sum_id$species, sum_id$site), main="Sample size by phase and species",
   legend = rownames(counts), beside=TRUE, las=2)
```

@@

## Predictors -->

### All predictors -->
```{r}
pca <- prcomp(na.omit(sum_site %>% 
              remove_rownames %>% 
              column_to_rownames(var="site") %>%
              select(lc_cover,ta_cover,ma_canopy,ta_canopy,rugosity,scar_bm,scar_den,tot_bm)),
  center = TRUE,scale. = TRUE)
summary(pca)

plot(pca,type="l")

fviz_pca_biplot(pca,
                col.var = "black",
                col.ind = as.factor(sum_site$island), # color by groups
                palette = c("#00AFBB", "#E7B800", "#FC4E07"),
                addEllipses = T, ellipse.level=0.95,
                ellipse.type = "confidence",
                label = "var",
                legend.title = "Groups",
                repel = T
                ) +
  scale_color_brewer(palette="Dark")+
  theme_minimal()

fviz_mca_biplot(pca, 
  habillage = sum_id$island, addEllipses = TRUE,
  label = "var", shape.var = 15) +
  scale_color_brewer(palette="Dark3")+
  theme_minimal()
```

<!-- ### Benthic variables -->
<!-- ```{r, echo=FALSE} -->
<!-- pred.benthic <- sum_site %>% select(lc_cover,ta_cover,ma_canopy,ta_canopy,rugosity) # should it matter if I use site level summary data vs. full follow dataset (where site-level data has been joined?) -->
<!-- pca1 <- prcomp(na.omit(pred.benthic),center = TRUE,scale. = TRUE)  -->

<!-- # add pc1 and pc2 as variables in dataset -->
<!-- pca1.scores <- as.data.frame(pca1$x) %>% select(pc1,pc2) -->
<!-- sum_site <- sum_site  %>% bind_cols(pca1.scores) -->
<!-- pca1.scores <- select(sum_site,site,pc1,pc2) -->
<!-- sum_id_pca <- sum_id %>% left_join(pca1.scores,by="site") -->

<!-- write.csv(sum_id_pca,file = here("data/sum_id_pca.csv")) -->
<!-- ``` -->
<!-- ### Fish community variables -->
<!-- ```{r} -->
<!-- pred.fish <- sum_site %>% select(scar_bm,scar_den,carn_bm) # should it matter if I use site level summary data vs. full follow dataset (where site-level data has been joined?) -->
<!-- pca2 <- prcomp(na.omit(pred.fish),center = TRUE,scale. = TRUE)  -->
<!-- summary(pca2) -->

<!-- plot(pca2,type="l") -->
<!-- ggbiplot(pca2, obs.scale = 1, var.scale = 1, groups=sum_site$island, ellipse = TRUE, circle = TRUE, loadings.label.repel=TRUE) + scale_color_manual(name="island", values=c("navy", "darkseagreen", "slategray2")) + theme(legend.direction = 'horizontal', legend.position = 'top') + theme_minimal() -->

<!-- ``` -->

<!-- ## Response variables -->

<!-- ### Grazing behaviors -->
<!-- *log transform increases P1 explained variance here...* -->
<!-- *Currently for_dur has several NA values (for individuals that didn't graze)... I am omitting that variable for now because I don't want to omit non-grazing individuals (instead would like to keep them as grazing=0* -->
<!-- ```{r} -->
<!-- response <- data %>% select(g_frac,tdb,br) -->
<!-- log.vars5 <- log(response[,1:3]) -->
<!-- pca3 <- prcomp(na.omit(response),center = TRUE,scale. = TRUE)  -->
<!-- summary(pca3) -->
<!-- plot(pca3,type="l") -->
<!-- ggbiplot(pca3, obs.scale = 1, var.scale = 1, groups=data$island,shape=data$island,ellipse = TRUE, circle = TRUE, alpha=0.8,varname.size = 5) + theme_minimal() + scale_color_manual(name="island", values=c("navy", "darkseagreen", "slategray2")) + labs(title="Grazing behaviors by island") + theme(plot.title = element_text(hjust = 0.5)) -->
<!-- ``` -->
<!-- ```{r} -->
<!-- response <- data %>% select(g_frac,fr,br) -->
<!-- log.vars5 <- log(response[,1:3]) -->
<!-- pca3 <- prcomp(na.omit(response),center = TRUE,scale. = TRUE)  -->
<!-- summary(pca3) -->
<!-- plot(pca3,type="l") -->
<!-- ggbiplot(pca3, obs.scale = 1, var.scale = 1,ellipse = TRUE, circle = TRUE, alpha=0.4,varname.size = 5) + theme(legend.direction = 'horizontal', legend.position = 'top') + theme_minimal() +geom_point(aes(col=data$scar_bm)) + labs(col="Scarid Biomass (g/100m2)", title="Grazing Behaviors by Scarid Biomass") + theme(plot.title = element_text(hjust = 0.5)) -->
<!-- ``` -->

<!-- ```{r,echo=FALSE} -->
<!-- vir <- data %>% filter(species_code=="stop") -->
<!-- response <- vir %>% select(g_frac,fr,br) -->
<!-- log.vars5 <- log(response[,1:3]) -->
<!-- pcavir <- prcomp(na.omit(response),center = TRUE,scale. = TRUE)  -->
<!-- ggbiplot(pcavir, obs.scale = 1, var.scale = 1,ellipse = TRUE, circle = TRUE, alpha=1,varname.size = 5) + theme(legend.direction = 'horizontal', legend.position = 'top') + theme_minimal() + labs(col="Scarid Biomass (g/100m2)", title="Grazing Behaviors by Scarid Biomass") + theme(plot.title = element_text(hjust = 0.5)) +geom_point(aes(colour=vir$scar_bm)) -->
<!-- ``` -->
<!-- ```{r,echo=FALSE} -->
<!-- vet <- data %>% filter(species_code=="qup") -->
<!-- response <- vet %>% select(g_frac,fr,br) -->
<!-- log.vars5 <- log(response[,1:3]) -->
<!-- pcavet <- prcomp(na.omit(response),center = TRUE,scale. = TRUE)  -->
<!-- ggbiplot(pcavet, obs.scale = 1, var.scale = 1,ellipse = TRUE, circle = TRUE, alpha=1,varname.size = 5) + theme(legend.direction = 'horizontal', legend.position = 'top') + theme_minimal() + labs(col="Scarid Biomass (g/100m2)", title="Grazing Behaviors by Scarid Biomass") + theme(plot.title = element_text(hjust = 0.5)) +geom_point(aes(colour=vet$scar_bm)) -->
<!-- ``` -->

# Model trials

@@

## Basic lm (to assess predictors of interest)
```{r}
initial<-data%>%filter(phase=="i") # for analysis with initial only
lm_fr <- lm(fr~scar_bm*species+phase+carn_bm+length_cm+pc1+pc2, data=data)
summary(lm_fr) # predicting feeding rate of initial phase only, all species
#vif(lm_fr)
```
*Notes: nearly everything is significant? Significant interaction between scar_bm and species for Spar. viride. carn_bm and scar_bm are highly positively correlated (~0.86) - is it ok to include both?*
*R2 goes up slightly with initial only*

```{r}
lm_g_frac <- lm(g_frac~species*scar_bm+phase+scar_den+carn_bm+phase+length_cm+pc1+pc2, data=data)
summary(lm_g_frac)
```
*Notes: carn_bm and pc1 are strongest predictors of time spent grazing, scar_bm and phase are also significant. Slight significance of species, length, and species-scar_bm interaction*

@@

```{r}
lme_fr <- lme(fr~phase+length_cm+species*scar_bm+scar_den+carn_bm+pc1+pc2,data=data,random=~1|island, na.action = na.omit)
summary(lme_fr)
```
*Notes: nearly everything is significant except for pc2 and scar_den?*
```{r}
lme_g_frac <- lme(g_frac~phase+length_cm+species*scar_bm+scar_den+carn_bm+pc1+pc2,data=data,random=~1|island, na.action = na.omit)
summary(lme_g_frac)
```
```{r}
lme_fr_stop <- lme(fr~phase+length_cm+scar_bm+scar_den+carn_bm+pc1+pc2,data=stop,random=~1|island, na.action = na.omit)
summary(lme_fr_stop)
```
*Notes: carn_bm and length are signifigant predictors*
```{r}
lme_fr_qup <- lme(fr~phase+length_cm+scar_bm+scar_den+carn_bm+pc1+pc2,data=qup,random=~1|island, na.action = na.omit)
summary(lme_fr_qup)
```
*Notes: pc1 is strongest (magnitude) predictors. Also significant are phase, carn_bm and scar_den*
# Model trials: species run individually

## Feeding rates
```{r}
lme_fr_stop <- lme(fr~phase+length_cm+scar_bm+scar_den+carn_bm+pc1+pc2,data=stop,random=~1|island, na.action = na.omit)
summary(lme_fr_stop)
```
*Notes: carn_bm and length are signifigant predictors*

```{r}
lme_fr_qup <- lme(fr~phase+length_cm+scar_bm+scar_den+carn_bm+pc1+pc2,data=qup,random=~1|island, na.action = na.omit)
summary(lme_fr_qup)
```
*Notes: pc1 is strongest (magnitude) predictors. Also significant are phase, carn_bm and scar_den*

## GAMs
```{r, echo=FALSE}
gam<-gam(fr~species+phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=data, method="REML")
summary(gam)
plot(gam,pages=1)
```
```{r, echo=FALSE}
gamvet<-gam(fr~phase+s(length_cm)+s(scar_bm)+(carn_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=vet, method="REML")
summary(gamvet)
plot(gamvet,pages=1)
```
```{r, echo=FALSE}
vir<-data %>% filter(species_code=="stop" & site!="Snapper Point")
gamvir<-gam(fr~phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=vir,method="REML")
summary(gamvir)
plot(gamvir,pages=1)
```

@@

### Feeding rate (bites per hour)
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(fr~species+phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=data, random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(fr~phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=vet, random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(fr~phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=vir, random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```

Most of variation is explained by species - for individual species models R2 values are very low (~0.3)

### Fraction of time spent grazing
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(g_frac~species+phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=data, random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(g_frac~phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=vet, random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(g_frac~phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=vir, random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```

Low R2 for all trials predicting g_frac (<0.4)(gammvir)
```
------------------------
To Do as of Nov. 7
* ~~incorporate pcs~~
* boosted regression trees ecosphere 2017 adrians paper
* ~~simpler graphs with just herb density x axis - grazing behaviors y, facet by species~~
* ~~species*density as fixed effect~~
* species as random effect
* ~~break out species and run separately~~

>>>>

---
title: "Investigation of scarid data"
author: "Molly Wilson"
date: "Dec. 2018"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    number_sections: true

@@

```{r,message=FALSE,warning=FALSE, echo=FALSE}
# set up
library(startR)
library(here)
library(tidyverse)
#library(RColorBrewer)
library(corrplot)
library(ggplot2)

library(nlme)
library(gamm4)
```
```{r,message=FALSE,warning=FALSE, echo=FALSE}
sum_site <- read.csv(here("data","sum_site.csv"), stringsAsFactors=F) %>%
  select(-X)
sum_id <- read.csv(here("data","sum_id.csv"), stringsAsFactors=F) %>%
  select(-X)
```

@@

```{r,echo=FALSE}
sample <- group_by(sum_id, site, species, phase, island) %>% summarize(n=n())

ggplot(sum_id, aes(site)) +
        facet_grid(island~species) +
        geom_bar(aes(fill = phase), position = "dodge") +
        scale_fill_manual(values = c("gray", "slategray2")) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1))

# playing with plot options...
ggplot(data = sample, aes(x = site, y = n, fill = phase)) + 
    geom_bar(stat = "identity", width = 1, position = "dodge") +
    facet_wrap(~island, strip.position = "bottom", scales = "free_x") +
    theme(panel.spacing = unit(0, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

@@

## Predictors

```{r}
pca_pred <- prcomp(na.omit(sum_site %>% 
              remove_rownames %>% 
              column_to_rownames(var="site") %>%
              select(lc_cover,ta_cover,ma_canopy,ta_canopy,rugosity,scar_bm,scar_den,tot_bm)),
  center = T,scale = T)

summary(pca_pred)
plot(pca_pred,type="l")

autoplot(pca_pred, data = sum_site, colour = 'island',
         loadings = T, loadings.colour = 'gray',
         loadings.label = T, loadings.label.size = 3, loadings.label.colour = 'black', 
            loadings.label.repel = T) +
    scale_color_manual(name="island", values=c("navy", "darkseagreen", "slategray2")) + 
    theme(legend.direction = 'horizontal', legend.position = 'top') +
    theme_minimal()

# add PC1 and PC2 as variables in dataset
sum_site_pca1 <- sum_site %>% 
  bind_cols(as.data.frame(pca_pred$x) %>% 
           select(PC1,PC2)) %>%
  clean_names(case="old_janitor")
sum_id_pca1 <- sum_id %>% left_join(select(sum_site_pca1,site,pc1,pc2), by = "site")
write.csv(sum_id_pca1,file=here("data/sum_id_pca1.csv"))
```

## Response variables

```{r}
pca_resp <- prcomp(na.omit(sum_id %>% 
              select(g_frac,fr,br)), 
  center = T,scale = T)

summary(pca_resp)
plot(pca_resp,type="l")

autoplot(pca_resp, data = sum_id, colour = "island",
         loadings = T, loadings.colour = 'gray',
         loadings.label = T, loadings.label.size = 3, loadings.label.colour = 'black', 
            loadings.label.repel = T) +
    scale_color_manual(name="island", values=c("navy", "darkseagreen", "slategray2")) + 
    theme(legend.direction = 'horizontal', legend.position = 'top') +
    theme_minimal()

# autoplot doesn't like when I add it for_dur or for_bites (NA values?)
```
# Model trials

@@

## Basic lm 
```{r}
lm_fr <- lm(fr~species+phase+length_cm+pc1+pc2, data=filter(sum_id_pca1, species_code!="rbp"))
summary(lm_fr) 
```

@@

```{r}
lme_fr <- lme(fr~phase+length_cm+species+pc1+pc2,data=filter(sum_id_pca1, species_code!="rbp"),random=~1|island, na.action = na.omit)
summary(lme_fr)
```

## GAMs
```{r, echo=FALSE}
gam<-gam(fr~species+phase+s(length_cm)+s(scar_bm)+s(pc1)+s(pc2), family=gaussian(link=identity),sum_id_pca1=data, method="REML")
summary(gam)
plot(gam,pages=1)
```

@@

```{r,message=FALSE,warning=FALSE}
gamm<-gamm(fr~phase+s(length_cm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=filter(sum_id_pca1, species_code=="qup"), random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```
```{r,message=FALSE,warning=FALSE}
gamm<-gamm(fr~phase+s(length_cm)+s(pc1)+s(pc2), family=gaussian(link=identity),data=filter(sum_id_pca1, species_code=="stop"), random=list(island=~1),method="REML")
summary(gamm$gam)
plot(gamm$gam,pages=1)
```

```
------------------------
To Do as of Nov. 7
* ~~incorporate pcs~~
* boosted regression trees ecosphere 2017 adrians paper
* ~~simpler graphs with just herb density x axis - grazing behaviors y, facet by species~~
* ~~species*density as fixed effect~~
* species as random effect
* ~~break out species and run separately~~
